Template for indicator documentation

### 2023-09-07 trophic_levels-biomass_indicators.rmd
Change location of data

_Author and date:_
James Speed

June 2023

@@

@@

@@

@@

```{r}
#| label: Data import
#| cache: TRUE

@@

#Livestock data from 1939 and vilt from 1938
names(livestockrast)<- sub(1939,1938,names(livestockrast))
names(livestockrast)

@@

```{r}
#| label: Plotting basic data
#| cache: TRUE

@@

```{r}
#| label: PlotTotalHerbivoreBiomass
#| cache: TRUE
#Total herbivore biomass
#For some reason need to read the raster in again
livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))

@@

```{r}
#| label: Plot carnivore biomass
#| cache: TRUE
carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))

@@

```{r}
#| label: Calculating expected herbivore biomass
#| cache: TRUE
# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
# npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]

expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  ggtitle("Expected herbivore biomass")+
  geom_spatraster(data=expected_herbivore_biomass)+facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
#writeRaster(expected_herbivore_biomass,paste0(pData,"/Trophic_levels/expectedherbivorebiomass.tif"),overwrite=TRUE)

@@

```{r}
#| label: DifferenceHerbivoreBiomass
#| cache: TRUE
#expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
#livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
totalherbivorebiomass_npp<-totalherbivorebiomass[[names(totalherbivorebiomass) %in% c("1999","2009","2015")]]

@@

#writeRaster(diffherb,paste0(pData,"/Trophic_levels/HerbivoreBiomassDiffExpected.tif"),overwrite=TRUE)

@@

#livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))

@@

#writeRaster(expected_carnivore_biomass,paste0(pData,"/Trophic_levels/expectedcarnivorebiomass.tif"),overwrite=TRUE)

#writeRaster(diffcarn,paste0(pData,"/Trophic_levels/CarnivoreBiomassDiffExpected.tif"),overwrite=TRUE)

@@

#Forest
#AR50 type raster
artype50<-rast("/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/AR50_Norge.tif")

#artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")

@@

#Load biomass rasters
# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
# viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
# livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
# carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
# expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
# expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")

@@

### NEED TO CHECK IF THIS IS THE CORRECT LAYER OR GET THE CORRECT ONE FROM JAMES

@@

 #writeRaster(forest_diffcarn,paste0(pData,"/Trophic_levels/forest_carnivores_diffexpected.tif"),overwrite=TRUE)
 #writeRaster(forest_diffherb,paste0(pData,"/Trophic_levels/forest_herbivores_diffexpected.tif"),overwrite=TRUE)
```

@@

#forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
#forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")

>>>>

_Author and date:_
James D. M. Speed, Gunnar Austrheim, Erling J. Solberg

August 2023

@@

```{r, echo=F}
Ecosystem <- "All" # e.g. "Skog og fjell"
Egenskap  <- "Biomass distribution among trophic levels" # e.g. "Primærproduksjon"
ECT       <- "Structural state characteristic"
Contact   <- "James Speed" # e.g. "Anders Kolstad"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

@@

## About the underlying data

@@

### Representativity in time and space

The data spans 1907 to 2015, with data points roughly every 10 years. The whole of Norway is covered. Herbivore data is collated at the municipality level (with the exception of semi-domestic reindeer in Finnmark at county level) and carnivore data at the county level. Municipalities and counties valid in 2007 were used. NPP is at a pixel resolution.

### Original units

The biomass of the trophic levels is as kg km^-2.

### Temporal coverage

The data spans 1907 to 2015, with data points roughly every 10 years.

## Ecosystem characteristic

### Norwegain standard

This indicator is within the ecosystem characterstic "biomass distribution among trophic levels"

### SEEA EA (UN standard)

This indicator alines most closely with B2 Structural state characteristics (although arguments could be made also for B1 and B3).

## Collinearities with other indicators

Colinearlities are likely to exist with climatic variables and disturbance.

## Reference condition and values

### Reference condition

Reference values are taken from published global relationships between plants and vertebrate herbivores ([Fløjgaard et al. 2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N)).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values

Deviations from the reference (expected) biomass of consumer based on consumed trophic level biomass are expressed as a two-sided percentage. Good condition was defined as within +/- 40% of the reference value (i.e. that expected from global relationships). Poor condition as over +/- 40% of the reference level. The choice of 40% is arbitary.

## Uncertainties

Uncertainty around the estimates have not been estimated. However, there are potential uncertainties around the reported livestock and hunting statistics and the demographic models used to estimate metabolic biomass at the population level. See Austrheim et al. 2011 and Speed et al. 2015 for further details.

## References

Austrheim, G., Solberg, E. J., & Mysterud, A. (2011). Spatio‐temporal variation in large herbivore pressure in Norway during 1949‐1999: has decreased grazing by livestock been countered by increased browsing by cervids?. Wildlife Biology, 17(3), 286-298.
Fløjgaard, C., Pedersen, P. B. M., Sandom, C. J., Svenning, J. C., & Ejrnæs, R. (2022). Exploring a natural baseline for large‐herbivore biomass in ecological restoration. Journal of Applied Ecology, 59(1), 18-24.
Hatton, I. A., McCann, K. S., Fryxell, J. M., Davies, T. J., Smerlak, M., Sinclair, A. R., & Loreau, M. (2015). The predator-prey power law: Biomass scaling across terrestrial and aquatic biomes. Science, 349(6252), aac6284.
Sobocinski, A. (2022). Mapping large carnivores in Norway across 175 years of changing policy (Master's thesis, NTNU). https://hdl.handle.net/11250/3047580 
Speed, J. D. M., Austrheim, G., Kolstad, A. L., & Solberg, E. J. (2019). Long-term changes in northern large-herbivore communities reveal differential rewilding rates in space and time. PLoS One, 14(5), e0217166.

## Additional information about the data

The raw data is available [here](10.6084/m9.figshare.24015072)

## Analyses

@@

```{r, message=FALSE, warning=FALSE}
#| label: Data import
#| cache: TRUE

@@

#Livestock data from 1939 and vilt from 1938
names(livestockrast)<- sub(1939,1938,names(livestockrast))
#names(livestockrast)

@@

```{r, message=FALSE, warning=FALSE}
#| label: Plotting basic data
#| cache: TRUE

@@

```{r, message=FALSE, warning=FALSE}
#| label: PlotTotalHerbivoreBiomass
#| cache: TRUE
#Total herbivore biomass
#For some reason need to read the raster in again
livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))

@@

```{r, message=FALSE, warning=FALSE}
#| label: Plot carnivore biomass
#| cache: TRUE
carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))

@@

```{r, message=FALSE, warning=FALSE}
#| label: Calculating expected herbivore biomass
#| cache: TRUE
expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  ggtitle("Expected herbivore biomass")+
  geom_spatraster(data=expected_herbivore_biomass)+facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))

@@

```{r}
#| label: DifferenceHerbivoreBiomass
#| cache: TRUE
totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
totalherbivorebiomass_npp<-totalherbivorebiomass[[names(totalherbivorebiomass) %in% c("1999","2009","2015")]]

@@

@@

@@

@@

#Forest
#AR50 type raster
# options(timeout = max(1000, getOption("timeout")))
# filedownload<-"downloadAR50.zip"
# download.file("https://ntnu.box.com/shared/static/mfrqmuyx14gtf8u9b8xlzb391m0f9fhy.zip","downloadAR50.zip")
# unzip('downloadAR50.zip',exdir="AR50")

artype50<-rast(paste0(pData, "AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif"))

@@

@@

@@

@@

 #Simplify by county nr
norcounty<-norcounty_shp %>% 
  group_by(FylkeNr) %>%
 summarise(geometry = st_union(geometry))
